Latent Predicate Networks: Concept Learning with Probabilistic Context-Sensitive Grammars

### 2014-10-21 drafts/cfp/draft.tex
Edits and reworking of discussion.

An LPN is an PRCG with 1 unary predicate $S$, and $K$ densely
connected binary predicates $\{A_k\}_{k=1}^K$ (Figure
\ref{fig:architecture}). Specifically, for each latent predicate
$A_k$, we define rules such that $A_k(w_a,w_b)$ is true for each
possible pair of terminals, $w_a,w_b \in T$. $T$ may include the empty
string, but we do not allow latent predicates of the form
$A_k(\epsilon, \epsilon)$: see Section~\ref{sec:implementation}. We also define rules such that $A_k(XY,UV)$
is true for each possible pair of predicates, $A_i,A_j$, and every
possible ordering of the variables $X,Y,U$ and $V$ across $A_i$ and
$A_j$. Finally, we define $S(XY)$ to be the concatenation of the two
arguments of $A_1(X,Y)$.

@@

Inference in hierarchical Dirichlet processes in stochastic grammars
and stochastic logic programs (SLPs) has been an active area of
research in recent decades~\cite{DBLP:conf/emnlp/LiangPJK07,
  goldwater2006contextual, johnson2006adaptor, cussens2001parameter}.
Variational inference is a popular approach in this domain and the one
we adopt here. As explained in the next section, we implemented
inference by translating LPNs into PRISM programs and using the
built-in Variational Bayes Expectation-Maximization
algorithm~\cite{sato2008variational}.

@@

lpn(P-IN) :- msw(P,V), reduce(P-IN,V).
		\end{verbatim}
		\subcaption{}
		\label{fig:prism}
	\end{minipage}
	\caption{Possible encodings of the 2-copy language $\{ww\,|\, w \in \{a,b\}^+\}$ as (\subref{fig:grammar}) an LPN, (\subref{fig:prism}) a PRISM program.}
	\label{fig:copy}
\end{figure}

LPNs can be encoded as a restricted and tractable subclass of
SLPs. Specifically, we encode LPNs with PRISM to obtain built-in
predicates for probabilistic execution and Bayesian inference over
SLPs (Figure \ref{fig:copy}). PRISM is an appropriate choice here,
because it implements a previously related extension of the
variational Bayesian Expectation Maximization algorithm for PCFGs to
SLPs satisfying a number of restrictions
\cite{cussens2001parameter}. Our LPN encodings satisfy these
restrictions, in particular the need for a finite explanation graph
({\it i.e.} set of all possible parse trees) for any
utterance. Because we disallow $A_i([\,],[\,])$ as a valid clause,
every term in the body of a clause has shorter arguments than the
head, giving finite derivations and a finite explanation graph.

First, we define the distribution over each non-terminal symbol, $n
\in N$, using the {\tt values} predicate. For each $n$ we define the
values the symbol can take on ({\it i.e.} one for each rule), as well
as the probability of each value. For each rule in the LPN, we also
define a corresponding clause of the {\tt reduce} predicate to capture
the actual relations between the left-hand and right-hand side of each
rule. Sampling and parsing begins with a call to {\tt lpn(P-IN)},
which chooses some rule for the left-hand side via {\tt msw(P,V)} and
recursively reduces the input term using the chosen rule ({\tt
reduce(P-IN,V)}).

@@

@@

LPNs seems particularly well-suited to learning in domains where the
number of concepts is infinite. In domains like these, the only way to
understand each concept is to discover structure in the sentences that
refer to these concepts. While number is a prime example, others
include spatial relations, kinship and other social relations, and
temporal relations. Future work should explore how LPNs might explain
learning in these domains, as well as how more complex number
relations might be learned.

Many of the algorithms and formalisms used here were originally
developed for use in other closely related areas, including logic
programming and semantic parsing. Semantic parsing, in particular,
seems especially related to the challenge we face here of jointly
learning meaning and structure from
sentences~\cite{berant2013semantic, liang2013learning,
  kwiatkowski2010inducing}. Semantic parsing, however, tends to frame
this challenge slightly differently than we do here, namely, by asking
how utterances can be mapped to an explicit internal logical
language. By contrast, we focus here on systems where meaning and
structure seem inseparable. Understanding how these two approaches
relate and inform one another is an interesting and open question.

>>>>

An LPN is an PRCG with the following generic architecture: the
\emph{axiom} of an LPN is a unary predicate $S$, and it has $K$
densely connected binary predicates $\{A_k\}_{k=1}^K$ (Figure
\ref{fig:architecture}). For each latent predicate $A_k$, we define
rules such that $A_k(w_a,w_b)$ is true for each possible pair of
terminals, $w_a,w_b \in T$. $T$ may include the empty string $\epsilon$, but we
do not allow latent predicates of the form $A_k(\epsilon, \epsilon)$:
see Section~\ref{sec:implementation}. We also define rules such that
$A_k(XY,UV)$ is true for each possible pair of predicates, $A_i,A_j$,
and every possible ordering of the variables $X,Y,U$ and $V$ across
$A_i$ and $A_j$. Finally, we define $S(XY)$ to be the concatenation of
the two arguments of $A_1(X,Y)$.

@@

Bayesian inference over stochastic grammars
and stochastic logic programs (CITE MUGGLETON) has been an active area of
research in recent decades~\cite{DBLP:conf/emnlp/LiangPJK07,
  goldwater2006contextual, johnson2006adaptor, cussens2001parameter}.
Variational inference is a popular approach in this domain and the one
we adopt here. As explained in the next section, we implemented
inference by translating LPNs into PRISM programs and using the
built-in Variational Bayes Expectation-Maximization
algorithm~\cite{sato2008variational}.

@@

lpn(P-IN) :- reduce(P-IN,V), msw(P,V).
		\end{verbatim}
		\subcaption{}
		\label{fig:prism}
	\end{minipage}
	\caption{Possible encodings of the 2-copy language $\{ww\,|\, w \in \{a,b\}^+\}$ as (\subref{fig:grammar}) an LPN, (\subref{fig:prism}) a PRISM program.}
	\label{fig:copy}
\end{figure}

LPNs can be encoded as a restricted subclass of PRISM programs; this
is very similar to how PCFGs are encoded in
PRISM~\cite{DBLP:conf/cl/2000}. A general probabilistic logic
programming system based on PROLOG, PRISM proves built-in predicates
for probabilistic execution and Bayesian inference over logic programs
with stochastic choices. There are several
restrictions placed on PRISM programs to maintain the validity of
their probabilistic interpretation. Most importantly, derivations of a
probabilistic predicate cannot contain cycles. Because we disallow
$A_i(\epsilon,\epsilon)$ as a valid clause, every term in the body of a clause
has shorter arguments than the head, giving acyclic derivations.

% First, we define the distribution over each non-terminal symbol, $n
% \in N$, using the {\tt values} predicate. For each $n$ we define the
% values the symbol can take on ({\it i.e.} one for each rule), as well
% as the probability of each value. For each rule in the LPN, we also
% define a corresponding clause of the {\tt reduce} predicate to capture
% the actual relations between the left-hand and right-hand side of each
% rule. Sampling and parsing begins with a call to {\tt lpn(P-IN)},
% which chooses some rule for the left-hand side via {\tt msw(P,V)} and
% recursively reduces the input term using the chosen rule ({\tt
% reduce(P-IN,V)}).

@@

Our LPN does not learn to how to transition between the last word in a decade
and the next decade (e.g. ``seventy nine'' to ``eighty''). Instead, it
makes the intuitively reasonable generalization that ``seventy nine''
should be followed by ``seventy ten.'' It will be interesting to see
how common such mistakes are in children learning to count.

@@

Though simple, the knowledge learned by LPNs in the simple number
domain we explore in this paper goes beyond the expressive capacity of
PCFGs and HMMs, and yet it avoids intractability by using a restricted
formalism. The human ability to learn abstract concepts must also rely
on such a compromise, and it remains to be seen to what extent LPNs
can model this ability.

This work leaves many open questions. What domains of concepts are
amenable to being represented and learned by LPNs? Can multiple and
varied systems of concepts be learned at once? Perhaps most
importantly, are there generic architectures for LPNs that allow a
larger number of predicates without an insurmountable blow-up in the
number of rules? One idea which we are pursuing is using layered
connectivity, much like neural networks, to limit the number of rules
in which each predicate participates.

Many of the algorithms and formalisms used here were originally
developed for use in other closely related areas, including logic
programming and semantic parsing. Semantic parsing, in particular,
seems especially related to the challenge we face here of jointly
learning meaning and structure from
sentences~\cite{berant2013semantic, liang2013learning,
  kwiatkowski2010inducing}. Semantic parsing, however, tends to frame
this challenge slightly differently than we do here, by asking
how utterances can be mapped to an explicit internal logical
language. By contrast, we focus here on systems where meaning and
structure seem inseparable. Understanding how these two approaches
relate and inform one another is an interesting and open question.
